Topological Equivariant Artist Model for Visualization Library Architecture

### 2021-10-27 paper/main.tex
k caption (borked) but fixes latex
shared nu

\IEEEPARstart{T}his paper uses methods from topology and category theory to develop a model of the transformation from data to graphical representation. This model provides a language to specify how data is structured and how this structure is carried through in the visualization, and serves as the basis for a functional approach to implementing visualization library components. Topology allows us to describe the structure of the data and graphics in a generalizable, scalable, and trackable way. Category theory provides a framework for separating the transformations implemented by visualization libraries from the various stages of visualization and therefore can be used to describe the constraints imposed on the library components \cite{wielsManagementEvolvingSpecifications1998,goguenCategoricalManifesto1991}. Well constrained modular components are inherently functional\cite{hughesWhyFunctionalProgramming1989}, and a functional framework yields a library implementation that is likely to be shorter, clearer, and more suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}. Using this functional approach, this paper contributes a practical framework for decoupling data processing from visualization generation in a way that allows for modular visualization components that are applicable to a variety of data sets in different formats. \note{is it OK that this is something reviewer 4 wrote}

@@

\subsection{Shared Index}
\begin{figure}[!h]
  \centering
  \subfloat[$\dbase_c = \dbase_a \cap \dbase_b$]{\includegraphics[width=.4\columnwidth]{union_k.png}%
  \label{fig:artist:union:k}}
  %\hfill
  \subfloat[$\dbase_c = \varnothing$]{\includegraphics[width=.4\columnwidth]{disjoint_k.png}%
  \label{fig:artist:union:disjoint_k}}
  \caption{In \autoref{fig:artist:union:k}, the input object $\sheaf_{\dtotal}$ encodes a continuous function over spaces $\dtotal_a$, $\dtotal_b$ 
  and overlapping space $\dtotal_c$. In this figure, the visual transformation is to a point on the line on screen. An artist that preserves the continuity of the function must generate graphics such that functions evaluted  $\dbasepoint \in \dbase_a \cap \dbase_b$ 
\end{figure}

@@

\begin{equation}
\label{eq:artist:union:k}
\begin{tikzcd}
  \dbase_{c} \arrow[d, "\iota", hook] \arrow[r, "\iota", hook] & \dbase_{b} \arrow[d, "i_{\dbase_b}"]      \\
  \dbase_{a} \arrow[r, "i_{\dbase_a}"]                         & \dbase_{a} \bigsqcup_{\dbase_{c}}\dbase_{b}
  \end{tikzcd}
\end{equation}
where the projection functions $i_{\dbase_a},i_{\dbase_b}$ behave such that $i_{\dbase_a}(\dbasepoint)|_{\dbasepoint \in \dbase_b} = \left[\dbasepoint\right]$, $i_{\dbase_b}(\dbasepoint)|_{\dbasepoint \in \dbase_b} = \left[\dbasepoint\right]$, meaning the projection functions yield equivalent points in the disjoint union $\dbase_a \bigsqcup_{\dbase_c} \dbase_b$. Data that is completely disjoint in the index, as shown in \autoref{fig:artist:union:disjoint_k}, is a special case where $\dbase_c=\varnothing$ and therefore there is no constraint on the artist with regards to how this data is rendered.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{intersection_f.png}
  \caption{The fiber spaces $F_a, F_b$ have a shared fiber space $F_c$. If they are input into the same artist $A_1$, then an equivariant transformation is one where $F_c$ is transformed to a visual element in a consistent matter. In this figure, $F_c$ is mapped to the x-axis for both $F_a$ and $F_b$.}
  \label{fig:artist:compose:union_fiber}
\end{figure}
an example of expecting consistency in overlapping fields are the multiple view constraints described by Qu and Hullman \cite{hullmanKeeping2018}.

@@

We enforce the equivariance constraint

\begin{equation}
  \label{eq:math:artist:nu_commute}
\begin{tikzcd}
  \dtotal_i \arrow[r] \arrow[r, "\vchannel_i"] \arrow[d, "m_{\delement}"'] & \vtotal_i \arrow[d, "m_{\velement}"] \\
  \dtotal_i \arrow[r, "\vchannel_i"]                           & \vtotal_i               
\end{tikzcd}
\end{equation}

One advantage of expressing the data to visual encoding $\vchannel_i$ as a functor from data fiber $\dfiber_i$ to $\vfiber_i$ is that it lets us formally express that encodings are bound to data and therefore should be consistent across views, as proposed by Hullamn and Qu\cite{hullmanKeeping2018}.

>>>>

\IEEEPARstart{T}his paper uses methods from algebraic topology and category theory to develop a model of the transformation from data to graphical representation. This model provides a language to specify how data is structured and how this structure is carried through in the visualization, and serves as the basis for a functional approach to implementing visualization library components. Topology allows us to describe the structure of the data and graphics in a generalizable, scalable, and trackable way. Category theory provides a framework for separating the transformations implemented by visualization libraries from the various stages of visualization and therefore can be used to describe the constraints imposed on the library components \cite{wielsManagementEvolvingSpecifications1998,goguenCategoricalManifesto1991}. Well constrained modular components are inherently functional\cite{hughesWhyFunctionalProgramming1989}, and a functional framework yields a library implementation that is likely to be shorter, clearer, and more suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}. Using this functional approach, this paper contributes a practical framework for decoupling data processing from visualization generation in a way that allows for modular visualization components that are applicable to a variety of data sets in different formats. \note{is it OK that this is something reviewer 4 wrote}

@@

\subsection{Shared Index}
\begin{figure}[!h]
  \centering
  \subfloat[$\dbase_c = \dbase_a \cap \dbase_b$]{\includegraphics[width=.4\columnwidth]{union_k.png}%
  \label{fig:artist:union:k}}
  %\hfill
  \subfloat[$\dbase_c = \varnothing$]{\includegraphics[width=.4\columnwidth]{disjoint_k.png}%
  \label{fig:artist:union:disjoint_k}}
  \caption{In \autoref{fig:artist:union:k}, the input object $\sheaf_{\dtotal}$ encodes a continuous function over spaces $\dtotal_a$, $\dtotal_b$ 
  and overlapping space $\dtotal_c$. In this figure, the visual transformation is to a point on the line on screen. An artist that preserves the continuity of the function must generate graphics such that functions evaluted  $\dbasepoint \in \dbase_a \cap \dbase_b$}
\end{figure}

@@

\begin{equation}
\label{eq:artist:intersection:basespace}
\begin{tikzcd}
  \dfiber_a \arrow[rrrr, "\vartist_a"] \arrow[rrd, "proj_a" description] &  &             &  & {\Gamma(\gbase_a, \gfiber_a)} \arrow[lld, "ga" description, dashed]  \\
                                                                         &  & \dfiber_{c} &  &                                                                      \\
  \dfiber_b \arrow[rrrr, "\vartist_b"] \arrow[rru, "proj_b" description] &  &             &  & {\Gamma(\gbase_b, \gfiber_b)} \arrow[llu, "g_b" description, dashed]
  \end{tikzcd}
\end{equation}
where the projection functions $i_{\dbase_a},i_{\dbase_b}$ behave such that $i_{\dbase_a}(\dbasepoint)|_{\dbasepoint \in \dbase_b} = \left[\dbasepoint\right]$, $i_{\dbase_b}(\dbasepoint)|_{\dbasepoint \in \dbase_b} = \left[\dbasepoint\right]$, meaning the projection functions yield equivalent points in the disjoint union $\dbase_a \bigsqcup_{\dbase_c} \dbase_b$. Data that is completely disjoint in the index, as shown in \autoref{fig:artist:union:disjoint_k}, is a special case where $\dbase_c=\varnothing$ and therefore there is no constraint on the artist with regards to how this data is rendered.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{intersection_f.png}
  \caption{The fiber spaces $F_a, F_b$ have a shared fiber space $F_c$. If they are input into the same artist $A_1$, then an equivariant transformation is one where $F_c$ is transformed to a visual element in a consistent matter. In this figure, $F_c$ is mapped to the x-axis for both $F_a$ and $F_b$.}
  \label{fig:artist:compose:union_fiber}
\end{figure}
an example of expecting consistency in overlapping fields are the multiple view constraints described by Qu and Hullman \cite{hullmanKeeping2018}. 
\begin{equation}
  \label{eq:artist:intersection:fiber}
  \begin{tikzcd}
    \dfiber_a \arrow[rrd, "proj_a" description] \arrow[rrrr, "\vartist_a"] &  &             &  & {\Gamma(\gbase_a, \gfiber_a)} \arrow[lld, "g_a" description, dashed] \\
                                                                           &  & \dfiber_{c} &  &                                                                      \\
    \dfiber_b \arrow[rru, "proj_b" description] \arrow[rrrr, "\vartist_b"] &  &             &  & {\Gamma(\gbase_a, \gfiber_a)} \arrow[llu, "g_b" description, dashed]
\end{tikzcd}
\end{equation}

% introduce notion of equivariant fiber, chasing diagram of F is different Gamma - Kosara and Zeminiskwy differentiable glyph 
% case 1: F3 \subset F1, {JFK} \in {JFK, EWR}
% case 2: F1 = F3 \times F1, F2= F3 \times F2, 
  % F1 ->F3 is projection -> Fiber consistency
% in that case, we can form the plus operator, disjoint union of A1 + A2 going in is consistent on subspace

We can compose these diagrams together to generate an \vartist\ to handle any of the things.

@@

We enforce the equivariance constraint specified in \autoref{eq:artist:construction:multi_nu} by constructing artists that apply \vchannel\ encoders such that 
\begin{equation}
  \label{eq:artist:construction:multi_nu}
  \begin{tikzcd}
    \dfiber_a \arrow[rrd, "proj_a" description] \arrow[rrrr, "\vchannel^{a}_{\dfiber_c}"] &  &                                                                                                                       &  & \vfiber^{a}_{\dfiber_c} \\
                                                                                          &  & \dfiber_{c} \arrow[rru, "\vchannel^{a}_{\dfiber_c}" description] \arrow[rrd, "\vchannel^{b}_{\dfiber_c}" description] &  &                         \\
    \dfiber_b \arrow[rru, "proj_b" description] \arrow[rrrr, "\vchannel^{a}_{\dfiber_c}"] &  &                                                                                                                       &  & \vfiber^{b}_{\dfiber_c}
    \end{tikzcd}
\end{equation}

the same fiber in two different data sets run through the same $\vchannel_{i}$ encoder will output to the same set of visual values in the corresponding visual fiber $\vfiber_i$.
